---
title: "Descriptive Analysis"
date: "4/22/2017"
output: pdf_document
---

# Teams Winning the Tournament

Which 5 teams have won the most NCAA titles?

```{r}
library(plyr)
library(dplyr)

# Get the champion for all years
tourney <- read.csv('./data/TourneyCompactResults.csv')
teams <- read.csv('./data/Teams.csv')
titleTeams <- merge(tourney, teams, by.x="Wteam", by.y="Team_Id")
titleTeams <- titleTeams %>% filter(Daynum == 154)
titleTeams <- titleTeams[, c('Team_Name', 'Season')] %>% arrange(desc(Season))

# Count the frequency of winning teams from 1985-2017
titleCount <- count(titleTeams, 'Team_Name') %>% arrange(desc(freq))
```

| Team           | Titles |
|----------------|--------|
| Duke           | 5      |
| Connecticut    | 4      |
| North Carolina | 4      |
| Kentucky       | 3      |
| Florida        | 2      |

```{r, echo=FALSE}
library(ggplot2)

p <- ggplot(head(titleCount, 5), aes(Team_Name, freq))
p + labs(title="Most Titles (1985-2017)", x="Team", y="Count") +
  geom_bar(stat = "identity")
```

# Teams in the Finals

Which teams have played in the championship game the most times?

```{r}
# We have the teams that have won the title, now we want teams that have lost in the finals.
finalsLosingTeams <- merge(tourney, teams, by.x="Lteam", by.y="Team_Id")
finalsLosingTeams <- finalsLosingTeams %>% filter(Daynum == 154)
finalsLosingTeams <- finalsLosingTeams[, c('Team_Name', 'Season')] %>% arrange(desc(Season))

# Combine all the winning teams with all the losing teams in the finals
finalsAppearances <- rbind(titleTeams, finalsLosingTeams)

# Count the frequency of winning teams from 1985-2017
finalsAppearancesCount <- count(finalsAppearances, 'Team_Name') %>% arrange(desc(freq))

```

| Team           | Finals Appearances |
|----------------|--------------------|
| Duke           | 9                  |
| North Carolina | 5                  |
| Kansas         | 5                  |
| Kentucky       | 5                  |
| Connecticut    | 4                  |

```{r, echo=FALSE}
library(ggplot2)

p <- ggplot(head(finalsAppearancesCount, 5), aes(Team_Name, freq))
p + labs(title="Most Finals Appearances (1985-2017)", x="Team", y="Count") + geom_bar(stat = "identity")
```

# Teams in the Final Four

How about the 5 teams that have appeared in the Final Four the most times?

```{r}
finalFourTeams <- tourney %>% filter(Daynum == 145 | Daynum == 146)
finalFourWinningTeams <- merge(finalFourTeams, teams, by.x="Wteam", by.y="Team_Id")
finalFourLosingTeams <- merge(finalFourTeams, teams, by.x="Lteam", by.y="Team_Id")
finalFourTeams <- rbind(finalFourWinningTeams, finalFourLosingTeams)
finalFourTeams <- finalFourTeams[, c('Team_Name', 'Season')] %>% arrange(desc(Season))

# Count the frequency
finalFourCount <- count(finalFourTeams, 'Team_Name') %>% arrange(desc(freq))
```

| Team           | Appearances |
|----------------|-------------|
| North Carolina | 17          |
| Kentucky       | 16          |
| Kansas         | 14          |
| Duke           | 14          |
| Arizona        | 10          |

```{r, echo=FALSE}
p <- ggplot(head(finalFourCount, 5), aes(Team_Name, freq))
p + labs(title="Most Final Four Appearances (1985-2017)", x="Team", y="Count") + geom_bar(stat = "identity")
```

# Significance of Statistics over Time

How have TOPG and PCT for the winning team from 2003-2017 changed over time?

```{r}
# All team season stats from 1985-2017
seasonStats <- read.csv('./data/FinalStats.csv')[,2:15]

# All championship teams from 1985-2017 with Team_Id column
titleTeams <- tourney %>% filter(Daynum == 154)
titleTeams <- titleTeams[, c('Wteam', 'Season')] %>% arrange(desc(Season))

# Function to return the season stats for each title team
getTeamStats <- function(seasonStats, teams) {
  datalist = list()

  for (i in 1:nrow(teams)) {
    datalist[[i]] <- subset(seasonStats, Season == teams[i, 'Season'] & TeamID == teams[i, 'Wteam'])
  }
  result <- bind_rows(datalist)
  result <- merge(result, teams, by.x="TeamID", by.y="Team_Id")
  result <- result[, !(names(result) %in% c('TeamID'))] %>% arrange(desc(Season))
  return(result)
}

# Title team season stats for years 2003-2017
titleTeamStats <- getTeamStats(seasonStats, titleTeams)
```

```{r, echo=FALSE}
p1 <- ggplot(data=titleTeamStats, aes(titleTeamStats['Season'])) +
  labs(title="Title Team Season Stats (2003-2017)", x="Season", y="Stat") + 
  geom_line(aes(y=titleTeamStats['TOPG'], color='Turnovers per game')) +
  geom_line(aes(y=titleTeamStats['ORPG'] + titleTeamStats['DRPG'], color='Rebounds per game')) +
  geom_line(aes(y=titleTeamStats['ORPG'] + titleTeamStats['APG'], color='Assists per game')) +
  geom_line(aes(y=titleTeamStats['ORPG'] + titleTeamStats['PFPG'], color='Personal fouls per game')) +
  geom_line(aes(y=titleTeamStats['ORPG'] + titleTeamStats['BPG'], color='Blocks per game')) +
  geom_line(aes(y=titleTeamStats['SPG'] + titleTeamStats['SPG'], color='Steals per game')) +
  geom_line(aes(y=titleTeamStats['PPG'], color='Points per game')) +
  scale_x_continuous(breaks=seq(2003, 2017, 2)) +
  scale_y_continuous(limits=c(0, 90))
p1

library(scales)

p2 <- ggplot(data=titleTeamStats, aes(titleTeamStats['Season'])) +
  labs(title="Title Team Season Percentage-based Stats (2003-2017)", x="Season", y="Stat") + 
  geom_line(aes(y=titleTeamStats['FGP'], color='Field goal %')) +
  geom_line(aes(y=titleTeamStats['TPP'], color='3 pt. %')) +
  geom_line(aes(y=titleTeamStats['PCT'], color='Win %')) +
  scale_x_continuous(breaks=seq(2003, 2017, 2)) +
  scale_y_continuous(labels = percent_format(), limits=c(0,1))
p2
```

How about TOPG for all teams from 2003-2017?

```{r}
# All final four teams from 1985-2017 with Team_Id column
tourneyTeams <- tourney %>% filter(Daynum > 132)
tourneyTeams <- tourneyTeams[, c('Wteam', 'Season')] %>% arrange(desc(Season))

# Tournament team season stats for years 2003-2017
tourneyTeamStats <- getTeamStats(seasonStats, tourneyTeams)

# Plot with regression line
p3 <- ggplot(data=tourneyTeamStats, aes(x=tourneyTeamStats['Season'], y=tourneyTeamStats['TOPG'])) +
  labs(title="Tournament Team Reg. Season Turnovers per Game by Year (2003-2017)", x="Season", y="%") + 
  geom_point(shape=1) + scale_x_continuous(breaks=seq(2003, 2017, 2)) +
  scale_y_continuous() + geom_smooth(method=lm)
p3
```

What about margin of victory over the years? Could the tournament be getting more competitive?

```{r}
scores <- tourney[, c('Wscore', 'Lscore', 'Season', 'Daynum')] %>% arrange(desc(Season))
scores <- scores %>% filter(Daynum > 144 & Daynum < 152)
scores <- mutate(scores, Margin = Wscore - Lscore)

avgScores <- group_by(scores, Season) %>% summarize(m = mean(Margin))

# Plot with regression line
p4 <- ggplot(data=avgScores, aes(x=avgScores['Season'], y=avgScores['m'])) +
  labs(title="Averaged Final Four Game Margins of Victory per Season (1985-2017)", x="Season", y="Margin of Victory (Points)") + 
  geom_point(shape=0) + scale_x_continuous(breaks=seq(1985, 2017, 2)) +
  scale_y_continuous() + geom_smooth(method=lm)
p4
```

What are the seed records in the first round?

```{r, echo=FALSE}
seeds <- read.csv('./data/TourneySeeds.csv')
seeds$Seed <- substring(seeds$Seed, 2, 3)

# Need to replace tourney$Wteam and tourney$Lteam with tourney$Seed
getTeamSeeds <- function(tournamentDataset, seedDataset) {
  tourneyCopy <- tournamentDataset
  colnames(tourneyCopy)[3] = 'Wseed'
  colnames(tourneyCopy)[5] = 'Lseed'

  tourneyCopy <- subset(tourneyCopy, Daynum < 138)
  
  for (i in 1:nrow(tournamentDataset)) {
    tourneySeason <- tournamentDataset[i, 1] # Season
    wteam <- tournamentDataset[i, 3] # Wteam
    lteam <- tournamentDataset[i, 5] # Lteam
    
    # Update Wteam with seed
    teamSeed <- as.numeric(subset(seeds, Season == tourneySeason & Team == wteam)$Seed)
    if (length(teamSeed) == 0) next
    tourneyCopy[i, 3] <- as.numeric(teamSeed)
    # Update Lteam with seed
    teamSeed <- as.numeric(subset(seeds, Season == tourneySeason & Team == lteam)$Seed)
    if (length(teamSeed) == 0) next
    tourneyCopy[i, 5] <- as.numeric(teamSeed)
  }
  return(tourneyCopy)
}

tourneyWithSeeds <- getTeamSeeds(tourney, seeds)

seedWinCount <- count(tourneyWithSeeds, 'Wseed') %>% arrange(Wseed)

p <- ggplot(seedWinCount)
p + labs(title="Wins in the first round (1985-2017)", x="Seed", y="Wins") +
  geom_bar(aes(x=Wseed, y=freq), stat = "identity", lwd=0.2) +
  scale_x_continuous(breaks=seq(1, 16, 1)) + geom_text(aes(x=Wseed, y=freq, label=freq), position=position_dodge(width=0.9), vjust=-0.25)
```
