---
title: "Descriptive Analysis"
date: "4/22/2017"
output: pdf_document
---

We want to create visualizations of the NCAA tournament over the years. Some ideas are:

- A graph of which 5 teams have won the most NCAA titles
- A graph of which 5 teams have entered the Final Four the most times
- A plot of TOPG and 3APG for the winning team from 2003-2017 (The idea is we want to see how these statistics have changed over time)

```{r}
library(dplyr)
library(plyr)

# Get the champion for all years
tourney <- read.csv('./data/TourneyCompactResults.csv')
teams <- read.csv('./data/Teams.csv')
tourney <- merge(tourney, teams, by.x="Wteam", by.y="Team_Id")
championshipGameTeams <- tourney %>% filter(Daynum == 154)
champions <- championshipGameTeams[, c('Team_Name', 'Season')] %>% arrange(desc(Season))

# Count the frequency of winning teams from 1985-2017
championCount <- count(champions, 'Team_Name') %>% arrange(desc(freq))
```

Create a bar plot of the championship teams from 1985-2017. 

```{r, echo=FALSE}
library(ggplot2)

p <- ggplot(head(championCount, 5), aes(Team_Name, freq))
p + labs(title="Most Titles (1985-2017)", x="Team", y="Count") + geom_bar(stat = "identity")
```
